>

### 2020-08-24 paper/response_Rjournal_reviewer1.txt

> Overall recommendation: accept, giving the authors the opportunity for final wordsmithing for clarity.

The paper has been has been revised including: simplifying sentance structure, minor changes for clarity, typos corrected. Some details comment in line below.

> I like the organization of the paper.  It seems like about the right amount of detail in terms of the background, discussion of the algorithm (with nice supporting figures), the basic package description, and examples.  The examples, in particular, are important and difficult for a paper of this type; I appreciate the care that went into the presentation.  The animations on the web were also appreciated.

Thank you, glad to hear the work is appreciated.

> I like the Github repository, and glanced through the package source.  It does conform to recommended package design/standards, including roxygen, etc…

Thanks for the brief review.

> I think the paper is generally well-written.  I found myself occasionally distracted by sentence structure. Some sentences are complex with qualifiers and additional phrases added with lots of commas (,).  As an example from the abstract, “This paper describes a new package, spinifex, to provide a manual tour, that allows the coefficient for a single variable to be controlled.”

> I can’t really argue it’s wrong, but it caused me to stop reading and think. I’d argue such a distraction doesn’t help.  Something like, “This paper describes a new package, spinifex, which provides manual tours of multivariate data by allowing the coefficient for a single variable to be controlled.” I can’t argue it’s a big deal and won’t argue that my proposal works or is that much better.  But as a reader, I’m interested in the subject matter and find it frustrating if I have to spend effort parsing a sentence.  Basically, any sentence with more than 3-4 commas could be studied.

> Here’s another: it really feels awkward to me.  I’d be inclined to use a comma after “identified” and do something else or eliminate the comma after “tour”.      “After a projection of interest is identified an analyst can then use the “finer brush” of the manual tour, by controlling the contributions of individual variables to explore the sensitivity they have on the structure visible in the projection. “

Ageed, these instances have been fixed and the whole paper has been reviewed with an eye out for sentance structure and over-use of commas.

> These are just examples.  Not a big deal, and frankly this is FAR better-written and presented than most papers I review: thank you!  Any further tightening before publication would help, but the subject and scholarship seems appropriate for publication in the R Journal.

Thank you for reviewing and offering the suggestions with examples. The feel as this round of edits did improve clarity, and simplfy sentances and clean up a few typos.

> Note a typo on the Github instructions: library(“spinifiex”) of course won’t work.

Corrected along with a few other typos.

### 2020-08-24 paper/spyrison-cook.Rmd
post proofing changes

---
title: > 
  spinifex: An R Package for Creating a Manual Tour of Low-dimensional  Projections of Multivariate Data
author:
  - name: Nicholas Spyrison
    affiliation: Monash University 
    address:
    - Faculty of Information Technology
    email:  nicholas.spyrison@monash.edu
  - name: Dianne Cook
    affiliation: Monash University 
    address:
    - Department of Econometrics and Business Statistics
    email:  dicook@monash.edu
abstract: >
  Dynamic low-dimensional linear projections of multivariate data collectively known as *tours* provide an important tool for exploring multivariate data and models. The R package \pkg{tourr} provides functions for several types of tours: grand, guided, little, local and frozen. Each of these can be viewed dynamically, or saved into a data object for animation. This paper describes a new package, \pkg{spinifex}, which provides a manual tour of multivariate data where the projection coefficient of a single variable is controlled. The variable is rotated fully into the projection, or completely out of the projection. The resulting sequence of projections can be displayed as an animation, with functions from either the \pkg{plotly} or \pkg{gganimate} packages. By varying the coefficient of a single variable, it is possible to explore the sensitivity of structure in the projection to that variable. This is particularly useful when used with a projection pursuit guided tour to simplify and understand the solution. The use of the manual tour is applied particle physics data to illustrate the sensitivity of structure in a projection to specific variable contributions.

keywords: exploratory data analysis, projection pursuit, high dimensional data, data visualization, statistical graphics, data science
preamble: >
  % preamble packages
link-citations: true
fig_caption: yes
toc: false
bibliography: spyrison-cook.bib
output: rticles::rjournal_article
  #bookdown::pdf_book:
   # base_format: rticles::rjournal_article #see sandbox for hard code solution.
---
\bibliography{spinifex_paper}
<!-- tourrd3 example has different setup and options and uses static images.  -->
```{r setup_paper, include=FALSE}
knitr::opts_chunk$set(
  fig.align = "center", 
  echo = TRUE, 
  collapse = TRUE,
  message = FALSE, 
  warning = FALSE, 
  error = FALSE,
  cache = TRUE, # needed for log and figure out?
  cache.lazy = FALSE
)
library(spinifex)
library(ggplot2)
library(gridExtra)
library(dplyr)
```

```{r paperFunctions, include=FALSE}
## FUNCTION FOR STATIC OUTPPUT, 
# going to facets loses control of output size, simpler to just live with this.
array2static <- function(.m_tour, .data, .m_var, .cat, .mag = 2.2){
  ##' for @EXAMPLE 
  if (F){ #Example run:
    set.seed(20190425); f_dat <- tourr::rescale(flea[,1:6]); f_cat <- factor(flea$species);
    f_path <- c(.693, -.022, .082, -.119, .706, .023, -.070, .438, .405, .515, .103, .604); 
    f_bas <- matrix(f_path, ncol=2); f_mvar <- 5; f_angle <- .29;
    f_mtour <- manual_tour(f_bas, manip_var = f_mvar, angle = f_angle);
    .m_tour <- f_mtour; .data <- f_dat; .m_var <- f_mvar; .cat <- f_cat; .mag = 2.2;
  }

  slides       <- array2df(array = .m_tour, data = .data)
  basis_slides <- slides$basis_slides
  data_slides  <- slides$data_slides
  n_slides     <- max(basis_slides$slide)
  if (n_slides != 15) 
    stop(paste0("stop: n_slides !=15 !!! Check the angle step size. n_slides = ", n_slides))

  # Initialize
  ## manip var asethetics
  p             <- nrow(basis_slides) / n_slides
  col_v         <- rep("grey80", p)
  col_v[.m_var] <- "blue"
  col_v         <- rep(col_v, n_slides)
  siz_v         <- rep(0.3, p)
  siz_v[.m_var] <- 1
  siz_v         <- rep(siz_v, n_slides)
  cat           <- rep(as.factor(.cat), n_slides)

  ## circle
  angle <- seq(0, 2 * pi, length = 180)
  circ  <- data.frame(c_x = cos(angle), c_y = sin(angle))
  circ[nrow(circ)+1, ] <- NA
  ## data asethetics
  data_slides <- data.frame(data_slides, cat = rep(.cat, n_slides))

  grid_b <- grid_t <-
    data.frame(slide = 1:n_slides, 
               x = .mag*rep(1:5, 3), y = .mag*rep(3:1, each = 5))
  grid_t$y <- grid_t$y + max(grid_t$y)
  # OUTER JOIN
  basis_grid <- merge(x = basis_slides, y = grid_t, by = "slide", all = TRUE)
  ## grid_t$x becomes x.y . ie. colname.tbl
  # CROSS JOIN
  circ_grid  <- merge(x = circ, y = grid_t, by = NULL)
  # OUTER JOIN
  data_grid  <- merge(x = data_slides, y = grid_b, by = "slide", all = TRUE)

  # BASIS AXES
  gg1 <-
    ggplot(data = basis_grid) +
    # AXES LINE SEGMETNS
    geom_segment(aes(x = x.x + x.y, y = y.x + y.y, xend = x.y, yend = y.y),
                 color = col_v, size = siz_v) +
    # AXES TEXT LABELS
    geom_text(aes(x = x.x + x.y, y = y.x + y.y, label = lab),
              color = col_v, vjust = "outward", hjust = "outward") +
    # AXES FRAME NUM
    geom_text(aes(x = x.y - .7, y = y.y + 1.1, 
                  label = paste0("frame: ",slide)), color = "grey50") +
    # AXES CIRCLE PATH
    suppressWarnings( # Suppress for "Removed 1 rows containing missing values."
      geom_path(data = circ_grid, color = "grey80",
                mapping = aes(x = x+c_x, y = y+c_y))
    )

  # PROJECTION
  gg2 <- gg1 +
    # PROJ DATA POINTS
    geom_point(data = data_grid, size = .7,
               mapping = aes(x = x.x + x.y, y = y.x + y.y, color = cat),
               shape = as.integer(cat) + 15) +
    # PROJ DATA FRAME NUM
    geom_text(data = data_grid, color = "grey50",
              mapping = aes(x = x.y - .7, y = y.y + 1.1, 
                            label = paste0("frame: ",slide))) +
    # OPTIONS
    theme_void() +
    scale_color_brewer(palette = "Dark2") + 
    theme(legend.position="none")
          #,panel.border = element_rect(colour = "black", fill = NA))

  gg2
}

view_basis2 <- function(basis, labels = paste0("V", 1:nrow(basis)),
                        manip_var, data = NULL, axes = "center"){
  basis <- as.data.frame(basis)
  angle <- seq(0, 2 * pi, length = 360)
  circ <- data.frame(x = cos(angle), y = sin(angle))
  zero <- set_axes_position(0, axes)
  basis <- set_axes_position(basis, axes)
  circ <- set_axes_position(circ, axes)
  col <- rep("black", nrow(basis))
  col[manip_var] <- "blue"

  gg <- ggplot2::ggplot() + ggplot2::scale_color_brewer(palette = "Dark2") + 
    ggplot2::theme_void() + ggplot2::theme(legend.position = "none") + 
    ggplot2::coord_fixed()
  if (axes != "off") {
    gg <- 
      gg + ggplot2::geom_path(
        data = circ, mapping = ggplot2::aes(x = x, y = y), 
        color = "grey80", size = 0.3, inherit.aes = F
      ) + 
      ggplot2::geom_segment(
        data = basis, color = col, 
        mapping = ggplot2::aes(x = V1, y = V2, xend = zero, yend = zero)
      ) + 
      ggplot2::geom_text(
        data = basis, mapping = ggplot2::aes(x = V1, y = V2, label = labels), 
        size = 4, color = col, hjust = 1.1, vjust=.4
      )
  }
  if (!is.null(data)) {
        proj <- as.data.frame(tourr::rescale(as.matrix(data) %*% 
            as.matrix(basis)) - 0.5)
        gg <- gg + ggplot2::geom_point(data = proj, 
                                       mapping = ggplot2::aes(x = V1, y = V2))
    }
    gg
}

view_manip_space2 <- function (basis,
                               manip_var,
                               manip_col = "blue",
                               tilt = pi *
                                 5 / 12,
                               z_col = "red",
                               labels = paste0("V", 1:nrow(basis)))
{
    find_angle <-
      function(a, b)
        acos(sum(a * b) / (sqrt(sum(a * a)) * sqrt(sum(b * b))) )
    make_curve <-
      function(ang_st = 0,
               ang_stop = 2 * pi) {
        # angle is atan(y/x)
        angle <- seq(ang_st, ang_stop,
                     length = round(360 / (2 * pi) * abs(ang_st - ang_stop)) )
        data.frame(x = cos(angle), y = sin(angle), z = sin(angle))
      }
    xyz <- function(x) { colnames(x)<-c("x","y","z");x }

    p <- nrow(basis)
    col_v <- rep("grey80", p)
    col_v[manip_var] <- manip_col
    siz_v <- rep(0.3, p)
    siz_v[manip_var] <- 1

    circ_r <- make_curve() %>%
      mutate( y=y*cos(tilt), z=z*sin(tilt))
    m_sp   <- xyz(as.data.frame(create_manip_space(basis, manip_var)))
    m_sp_r <- m_sp %>% mutate( y=y*cos(tilt), z=z*sin(tilt))
    mvar   <- m_sp[manip_var, ]
    mvar_r <- m_sp_r[manip_var, ]

    phi_ang     <- find_angle(c(m_sp_r$x, m_sp_r$y), c(m_sp_r$x, m_sp_r$z))
    phi_offset  <- find_angle(c(mvar_r$x, mvar_r$y), c(1,0))
    phi_curve_r <- .4*make_curve(ang_st = phi_offset, ang_stop = phi_ang) %>%
      mutate( y=y*cos(tilt), z=z*sin(tilt))

    thata_ang     <- find_angle(c(mvar$x, mvar$y),c(1, 0))
    theta_curve_r <- .5*make_curve(ang_st = 0, ang_stop = thata_ang) %>%
      mutate( y=y*cos(tilt), z=z*sin(tilt))

    gg <-
      ggplot2::ggplot() + ggplot2::scale_color_brewer(palette = "Dark2") +
      ggplot2::theme_void() + ggplot2::theme(legend.position = "none") +
      ggplot2::coord_fixed() +
        ggplot2::geom_path(data = circ_r,
                           mapping = ggplot2::aes(x = x, y = y),
                           color = manip_col,
                           size = 0.3,
                           inherit.aes = F
        ) +
        #projection plane:
        ggplot2::geom_segment(data = m_sp_r,
                              mapping = ggplot2::aes(
                                x = x,
                                y = y,
                                xend = 0,
                                yend = 0
                              ),
                              size = siz_v,
                              colour = col_v
        ) +
        ggplot2::geom_text(data = m_sp_r,
                           mapping = ggplot2::aes(x = x, y = y, label = labels),
                           size = 4,
                           colour = col_v,
                           vjust = "outward",
                           hjust = "outward"
        ) +
        #Z direction
        ggplot2::geom_path(data = circ_r,
                           mapping = ggplot2::aes(x = x, y = z),
                           color = z_col,
                           size = 0.3,
                           inherit.aes = F
        ) +
        ggplot2::geom_segment(data = mvar_r,
                              mapping = ggplot2::aes(x = x,y = z,
                                                     xend = 0,yend = 0),
                              size = 1,
                              colour = z_col
        ) +
        ggplot2::geom_segment(data = mvar_r,
                              mapping = ggplot2::aes(x = x, y = z,
                                                     xend = x,yend = y),
                              size = 0.3,
                              colour = "grey80",
                              linetype = 2
        ) +
        ggplot2::geom_text(data = mvar_r,
                           mapping = ggplot2::aes(x = x, y = z, label = labels[manip_var]),
                           size = 4,
                           colour = z_col,
                           vjust = "outward",
                           hjust = "outward"
        ) +
        # label angles
        geom_path(data = phi_curve_r,
                  mapping = aes(x=x, y=z),
                  color = z_col,
                  size = 0.2
        ) +
        geom_text(data = 1.2 * phi_curve_r[ceiling(nrow(phi_curve_r)/2), ],
                  mapping = ggplot2::aes(x=x, y=z, label = "phi"),
                  color = z_col,
                  parse = T,
                  size = 4
        ) +
        geom_path(data = theta_curve_r,
                  mapping = aes(x, y),
                  color = manip_col,
                  size = 0.2
        ) +
        geom_text(data = 1.2 * theta_curve_r[ceiling(nrow(theta_curve_r)/2), ],
                  mapping = ggplot2::aes(x=x, y=y-.02, label = "theta"),
                  color = manip_col,
                  parse = T,
                  size = 4
        )
    gg
  }
```

@@

<!-- Tours -->
Exploring multivariate spaces is a challenging task, increasingly so as dimensionality increases. Traditionally, static low-dimensional projections are used to display multivariate data in two dimensions including principal component analysis, linear discriminant spaces or projection pursuit. These are useful for finding relationships between multiple variables, but they are limited because they show only a glimpse of the high-dimensional space. An alternative approach is to use a tour [@asimov_grand_1985] of dynamic linear projections to look at many different low-dimensional projections. Tours can be considered to extend the dimensionality of visualization, which is important as data and models exist in more than 3D. The package \CRANpkg{tourr} [@wickham_tourr_2011] provides a platform for generating tours. It can produce a variety of tours, each paired with a variety of possible displays. A user can make a grand, guided, little, local or frozen tour, and display the resulting projected data as a scatterplot, density plot, histogram, or even as Chernoff faces if the projection dimension is more than 3.

@@

<!-- Paper outline -->
The paper is organized as follows. Section \ref{sec:algorithm} describes the algorithm used to perform a radial manual tour as implemented in the package \CRANpkg{spinifex}. Section \ref{sec:display} explains how to generate an animation of the manual tour using the animation frameworks offered by \CRANpkg{plotly} [@sievert_plotly_2018] and \CRANpkg{gganimate} [@pedersen_gganimate:_2019]. Package functionality and code usage following the order applied in the algorithm follows in section \ref{sec:usage}. Section \ref{sec:application} illustrates how this can be used for sensitivity analysis applied to multivariate data collected on high energy physics experiments [@wang_mapping_2018]. Section \ref{sec:discussion} summarizes this paper and discusses potential future directions.

@@

<!-- Projection basis -->
An initial 2D projection basis must be provided. A suggested way to start is to identify an interesting projection using a projection pursuit guided tour. Here the holes index is used to find a 2D projection of the flea data, which shows three separated species groups. Figure \ref{fig:step0} shows the initial projection of the data. The left panel displays the projection basis ($\textbf{B}$) and can be used as a visual guide of the magnitude and direction that each variable contributes to the projection. The right panel shows the projected data, $\textbf{Y}_{[n,~2]} ~=~ \textbf{X}_{[n,~p]} \textbf{B}_{[p,~2]}$. The color and shape of points are mapped to the flea species. This plot is made using the \code{view\_basis()} function in \pkg{spinifex}, which generates a \CRANpkg{ggplot2} [@wickham_ggplot2:_2016] object.

```{r step0, echo=F, results='hide', out.width='70%',                             fig.cap = "Initial 2D projection: representation of the basis  (left) and resulting data projection (right) of standardized flea data. The color and shape of data points are mapped to beetle species. The basis was identified using a projection pursuit guided tour, with the holes index. The contribution of the variables aede2 and tars1 approximately contrasts the other variables. The visible structure in the projection are the three clusters corresponding to the three species. Produced with the function \\code{view\\_basis()}."}
## Flea holes tour
set.seed(20190425)
f_dat  <- tourr::rescale(flea[,1:6])
f_cat  <- factor(flea$species)
## hard code a basis. tourr doesn't fix results with set.seed().
f_path <-  c(.693, -.022, .082, -.119, .706, .023, 
             -.070, .438, .405, .515, .103, .604)
f_bas <- matrix(f_path, ncol=2)
# f_path <- save_history(f_dat, guided_tour(holes()))
# f_bas  <- matrix(f_path[,, max(dim(f_path)[3])], ncol=2)
f_mvar <- 5
f_msp  <- create_manip_space(basis = f_bas, manip_var = f_mvar)
f_proj <- data.frame(tourr::rescale(f_dat %*% f_msp[, 1:2]))

# step0, output
# Adjust centering to make axes and data side by side
view_basis(f_bas, lab = colnames(f_dat)) +
  geom_point(data = f_proj,
             mapping = aes(x = X1 + .75, y = X2 - .5, color = f_cat),
             pch = as.integer(f_cat) + 15) +
  theme(panel.border = element_rect(colour = "black", fill = NA))
```

@@

```{r step2, echo=F, out.width='70%',                                           fig.cap = "Illustration of a 3D manip space, this space is rotated effectively changing the contribution of the manip variable, aede2 in the example data. The blue circle and variable map lies on the projection plane. The red circle, orthogonal to the projection plane, illustrates the manipulation space and how the manip var can be controlled and how this affects the variable contribution back onto the projection plane. The other variables are omitted from the manipulation dimension for simplicity. Picturing the other variables in that dimension reveals the intuition that rotating one variable performs a constrained rotation on the others. This is illustrated with the \\code{view\\_manip\\_space()} function."}
view_manip_space2(basis = f_bas, manip_var = f_mvar, labels = colnames(f_dat)) +
    theme(panel.border = element_rect(colour = "black", fill = NA))
```

@@

```{r step3, echo=F, warning=F, fig.height=2, fig.width=8, out.width='100%',  fig.cap = "Snapshots of a radial manual tour manipulating aede2: (1) original projection, (2) full contribution, (3) zero contribution, (4) back to original. "}
f_angle <- .29
f_mtour <- manual_tour(f_bas, manip_var = f_mvar, angle = f_angle)
if(dim(f_mtour)[3] != 15) message(
  paste0("!!! n_slides != 15 !!! Check angle step size. n_slides = ", dim(f_mtour)[3], ". angle = ", f_angle))
p1 <- view_basis2(f_mtour[,,1],  colnames(f_dat), manip_var = f_mvar) + ggtitle("1 phi=0.78")
p2 <- view_basis2(f_mtour[,,5],  colnames(f_dat), manip_var = f_mvar) + ggtitle("2  phi=0")
p3 <- view_basis2(f_mtour[,,12], colnames(f_dat), manip_var = f_mvar) + ggtitle("3 phi=1.57")
p4 <- view_basis2(f_mtour[,,15], colnames(f_dat), manip_var = f_mvar) + ggtitle("4 phi=0.78")
grid.arrange(p1, p2, p3, p4, ncol=4)
```

>>>>

---
title: > 
  spinifex: An R Package for Creating a Manual Tour of Low-dimensional Projections of Multivariate Data
author:
  - name: Nicholas Spyrison
    affiliation: Monash University 
    address:
    - Faculty of Information Technology
    email:  nicholas.spyrison@monash.edu
  - name: Dianne Cook
    affiliation: Monash University 
    address:
    - Department of Econometrics and Business Statistics
    email:  dicook@monash.edu
abstract: >
  Dynamic low-dimensional linear projections of multivariate data collectively known as *tours* provide an important tool for exploring multivariate data and models. The R package \pkg{tourr} provides functions for several types of tours: grand, guided, little, local and frozen. Each of these can be viewed dynamically, or saved into a data object for animation. This paper describes a new package, \pkg{spinifex}, which provides a manual tour of multivariate data where the projection coefficient of a single variable is controlled. The variable is rotated fully into the projection, or completely out of the projection. The resulting sequence of projections can be displayed as an animation, with functions from either the \pkg{plotly} or \pkg{gganimate} packages. By varying the coefficient of a single variable, it is possible to explore the sensitivity of structure in the projection to that variable. This is particularly useful when used with a projection pursuit guided tour to simplify and understand the solution. The use of the manual tour is applied particle physics data to illustrate the sensitivity of structure in a projection to specific variable contributions.
keywords: exploratory data analysis, projection pursuit, high dimensional data, data visualization, statistical graphics, data science
preamble: >
  % preamble packages
link-citations: true
fig_caption: yes
toc: false
bibliography: spyrison-cook.bib
output: rticles::rjournal_article
  #bookdown::pdf_book ## alternative
---
```{r setup_paper, include=FALSE}
knitr::opts_chunk$set(
  fig.align  = "center", 
  echo       = TRUE, 
  collapse   = TRUE,
  message    = FALSE, 
  warning    = FALSE, 
  error      = FALSE,
  cache      = TRUE, # needed for log and figure out?
  cache.lazy = FALSE
)
if (F){ ## From reproducing, please use this dev ver:
  remotes::install_github("nspyrison/spinifex", 
                          ref = "da75976588dc008e21f01ce6a57d9778d0ed6933")
}
library(tourr)
library(spinifex) ## Use dev version listed above!
library(ggplot2)
library(gridExtra)
library(dplyr)
```

```{r paperFunctions, include=FALSE}
## FUNCTION FOR STATIC OUTPPUT, 
# going to facets loses control of multiple geoms and output size, simpler to just live with this.
array2ggfacets <- function(tour_array, data, m_var, class, margin = 2.2){
  n_frames <- dim(tour_array)[3]
  if (n_frames != 15) 
    stop(paste0("stop: n_frames != 15!!! Check the angle step size. n_frames = ", n_frames))

  ## Initialize
  frames       <- array2df(array = tour_array, data = data)
  basis_frames <- frames$basis_frames
  data_frames  <- frames$data_frames
  p            <- nrow(basis_frames) / n_frames

  ## manip var asethetics
  col_v        <- rep("grey80", p)
  col_v[m_var] <- "blue"
  col_v        <- rep(col_v, n_frames)
  siz_v        <- rep(0.3, p)
  siz_v[m_var] <- 1
  siz_v        <- rep(siz_v, n_frames)
  cat          <- rep(as.factor(class), n_frames)

  ## circle
  angle <- seq(0, 2 * pi, length = 180)
  circ  <- data.frame(c_x = cos(angle), c_y = sin(angle))
  circ[nrow(circ)+1, ] <- NA
  ## Data asethetics
  data_frames <- data.frame(data_frames, class = rep(class, n_frames))
  grid_b <- grid_t <- data.frame(
    frame = 1:n_frames, x = margin * rep(1:5, 3), y = margin * rep(3:1, each = 5))
  grid_t$y <- grid_t$y + max(grid_t$y)
  ## OUTER JOIN
  basis_grid <- merge(x = basis_frames, y = grid_t, by = "frame", all = TRUE)
  ## CROSS JOIN
  circ_grid  <- merge(x = circ, y = grid_t, by = NULL)
  ## OUTER JOIN
  data_grid  <- merge(x = data_frames, y = grid_b, by = "frame", all = TRUE)

  ##### RENDER
  ## SETUP
  gg <- ggplot(data = basis_grid) +
    ## AXES LINE SEGMETNS
    geom_segment(aes(x = x.x + x.y, y = y.x + y.y, xend = x.y, yend = y.y),
                 color = col_v, size = siz_v) +
    ## AXES TEXT LABELS
    geom_text(aes(x = x.x + x.y, y = y.x + y.y, label = lab),
              color = col_v, vjust = "outward", hjust = "outward") +
    ## AXES FRAME NUM
    geom_text(aes(x = x.y - .7, y = y.y + 1.1, 
                  label = paste0("frame: ",frame)), color = "grey50") +
    ## AXES CIRCLE PATH
    suppressWarnings( # Suppress for "Removed 1 rows containing missing values."
      geom_path(data = circ_grid, color = "grey50",
                mapping = aes(x = x + c_x, y = y + c_y))
    )

  ## PROJECTION
  gg <- gg +
    ## PROJ DATA POINTS
    geom_point(data = data_grid, size = .7,
               mapping = aes(x = x.x + x.y, y = y.x + y.y, 
                             color = class, shape = class)) +
    ## FACET FRAME NUM
    geom_text(data = data_grid, color = "grey50",
              mapping = aes(x = x.y - .7, y = y.y + 1.1, 
                            label = paste0("frame: ",frame))) +
    theme_void() +
    scale_color_brewer(palette = "Dark2") + 
    theme(legend.position = "none")

  ## Return
  gg
}

view_manip_space2 <- function (basis,
                               manip_var,
                               manip_col = "blue",
                               tilt = pi * 5 / 12,
                               z_col = "red",
                               labels = paste0("V", 1:nrow(basis)))
{
  #### Finds the angle between two vectors
  find_angle <- function(a, b)
    acos(sum(a * b) / (sqrt(sum(a * a)) * sqrt(sum(b * b))) )
  #### Makes a df semi-circle, with 1 row per degree
  make_curve <- function(ang_st = 0,
                         ang_stop = 2 * pi) {
      degrees <- round(360 / (2 * pi) * abs(ang_st - ang_stop))
      angle <- seq(ang_st, ang_stop, length = degrees)
      data.frame(x = cos(angle), y = sin(angle), z = sin(angle))
    }

  ## Initialize
  p <- nrow(basis)
  m_sp   <- as.data.frame(create_manip_space(basis, manip_var))
  colnames(m_sp) <- c("x", "y", "z")
  m_sp_r <- m_sp %>% mutate(y = y * cos(tilt), z = z * sin(tilt))
  mvar   <- m_sp[manip_var, ]
  mvar_r <- m_sp_r[manip_var, ]

  col_v <- rep("grey80", p)
  col_v[manip_var] <- manip_col
  siz_v <- rep(0.3, p)
  siz_v[manip_var] <- 1

  circ_r <- make_curve() %>%
    mutate(y = y * cos(tilt), z = z * sin(tilt))

  phi_ang     <- find_angle(c(m_sp_r$x, m_sp_r$y), c(m_sp_r$x, m_sp_r$z))
  phi_offset  <- find_angle(c(mvar_r$x, mvar_r$y), c(1,0))
  phi_curve_r <- .4 * make_curve(ang_st = phi_offset, ang_stop = phi_ang) %>%
    mutate(y = y * cos(tilt), z = z * sin(tilt))

  thata_ang     <- find_angle(c(mvar$x, mvar$y),c(1, 0))
  theta_curve_r <- .5 * make_curve(ang_st = 0, ang_stop = thata_ang) %>%
    mutate( y=y*cos(tilt), z=z*sin(tilt))

  ## Render
  gg <-
    ggplot2::ggplot() + ggplot2::scale_color_brewer(palette = "Dark2") +
    ggplot2::theme_void() + ggplot2::theme(legend.position = "none") +
    ggplot2::coord_fixed() +
    ## Circle
    ggplot2::geom_path(data = circ_r,
                       mapping = ggplot2::aes(x = x, y = y),
                       color = manip_col, size = 0.3, inherit.aes = FALSE) +
    ## Projection plane:
    ggplot2::geom_segment(data = m_sp_r,
                          mapping = ggplot2::aes(
                            x = x,
                            y = y,
                            xend = 0,
                            yend = 0
                          ),
                          size = siz_v, colour = col_v) +
    ggplot2::geom_text(data = m_sp_r,
                       mapping = ggplot2::aes(x = x, y = y, label = labels),
                       size = 4, colour = col_v,
                       vjust = "outward", hjust = "outward") +
    ## Z direction
    ggplot2::geom_path(data = circ_r,
                       mapping = ggplot2::aes(x = x, y = z),
                       color = z_col, size = 0.3, inherit.aes = FALSE) +
    ggplot2::geom_segment(data = mvar_r,
                          mapping = ggplot2::aes(x = x,y = z,
                                                 xend = 0,yend = 0),
                          size = 1, colour = z_col) +
    ggplot2::geom_segment(data = mvar_r,
                          mapping = ggplot2::aes(x = x, y = z,
                                                 xend = x,yend = y),
                          size = 0.3, colour = "grey80", linetype = 2) +
    ggplot2::geom_text(data = mvar_r,
                       mapping = ggplot2::aes(x = x, y = z, 
                                              label = labels[manip_var]),
                       size = 4, colour = z_col,
                       vjust = "outward",
                       hjust = "outward"
    ) +
    ## Label angles
    geom_path(data = phi_curve_r,
              mapping = aes(x=x, y=z),
              color = z_col, size = 0.2) +
    geom_text(data = 1.2 * phi_curve_r[ceiling(nrow(phi_curve_r)/2), ],
              mapping = ggplot2::aes(x=x, y=z, label = "phi"),
              color = z_col, size = 4, parse = TRUE) +
    geom_path(data = theta_curve_r,
              mapping = aes(x, y),
              color = manip_col,
              size = 0.2) +
    geom_text(data = 1.2 * theta_curve_r[ceiling(nrow(theta_curve_r)/2), ],
              mapping = ggplot2::aes(x=x, y=y-.02, label = "theta"),
              color = z_col, size = 4, parse = TRUE)

  ## Return
  gg
}
```

@@

<!-- Tours -->
Exploring multivariate spaces is a challenging task, increasingly so as dimensionality increases. Traditionally, static low-dimensional projections are used to display multivariate data in two dimensions including principal component analysis, linear discriminant spaces or projection pursuit. These are useful for finding relationships between multiple variables, but they are limited because they show only a glimpse of the high-dimensional space. An alternative approach is to use a tour [@asimov_grand_1985] of dynamic linear projections to look at many different low-dimensional projections. Tours can be considered to extend the dimensionality of visualization, which is important as data and models exist in more than 3D. The package \CRANpkg{tourr} [@wickham_tourr:_2011] provides a platform for generating tours. It can produce a variety of tours, each paired with a variety of possible displays. A user can make a grand, guided, little, local or frozen tour, and display the resulting projected data as a scatterplot, density plot, histogram, or even as Chernoff faces if the projection dimension is more than 3.

@@

<!-- Paper outline -->
The paper is organized as follows. Section \ref{sec:algorithm} describes the algorithm used to perform a radial manual tour as implemented in the package \CRANpkg{spinifex}. Section \ref{sec:display} explains how to generate an animation of the manual tour using the animation frameworks offered by \CRANpkg{plotly} [@sievert_interactive_2020] and \CRANpkg{gganimate} [@pedersen_gganimate_2020]. Package functionality and code usage following the order applied in the algorithm follows in section \ref{sec:usage}. Section \ref{sec:application} illustrates how this can be used for sensitivity analysis applied to multivariate data collected on high energy physics experiments [@wang_mapping_2018]. Section \ref{sec:discussion} summarizes this paper and discusses potential future directions.

@@

<!-- Projection basis -->
An initial 2D projection basis must be provided. A suggested way to start is to identify an interesting projection using a projection pursuit guided tour. Here the holes index is used to find a 2D projection of the flea data, which shows three separated species groups. Figure \ref{fig:step0} shows the initial projection of the data. The left panel displays the projection basis ($\textbf{B}$) and can be used as a visual guide of the magnitude and direction that each variable contributes to the projection. The right panel shows the projected data, $\textbf{Y}_{[n,~2]} ~=~ \textbf{X}_{[n,~p]} \textbf{B}_{[p,~2]}$. The color and shape of points are mapped to the flea species. This plot is made using the \code{view\_basis()} function in \pkg{spinifex}, which generates a \CRANpkg{ggplot2} [@wickham_ggplot2_2016] object.

```{r step0, echo=F, results='hide', out.width='70%',                        fig.cap = "Initial 2D projection: representation of the basis (left) and resulting data projection (right) of standardized flea data. The color and shape of data points are mapped to beetle species. The basis was identified using a projection pursuit guided tour, with the holes index. The contribution of the variables aede2 and tars1 approximately contrasts the other variables. The visible structure in the projection are the three clusters corresponding to the three species. Produced with the function \\code{view\\_basis()}."}
## Flea holes tour
set.seed(20190425)
f_dat  <- tourr::rescale(flea[,1:6])
f_cat  <- factor(flea$species)
## hard code a basis. tourr doesn't fix results with set.seed().
f_bas <-  c(.693, -.022, .082, -.119, .706, .023, 
             -.070, .438, .405, .515, .103, .604) %>% 
  matrix( ncol=2) %>% 
  tourr::orthonormalise()
f_mvar <- 5
f_msp  <- create_manip_space(basis = f_bas, manip_var = f_mvar)
f_proj <- data.frame(tourr::rescale(f_dat %*% f_msp[, 1:2]))

# step0, output
# Adjust centering to make axes and data side by side
oblique_frame(f_bas, data = f_dat, manip_var = f_mvar, lab = colnames(f_dat),
              color = f_cat, shape = f_cat) + 
  coord_fixed() + theme_void() + 
  theme(legend.position = "none")
```

@@

```{r step2, echo=F, out.width='70%',                                                fig.cap = "Illustration of a 3D manip space, this space is rotated effectively changing the contribution of the manip variable, aede2 in the example data. The blue circle and variable map lies on the projection plane. The red circle, orthogonal to the projection plane, illustrates the manipulation space and how the manip var can be controlled and how this affects the variable contribution back onto the projection plane. The other variables are omitted from the manipulation dimension for simplicity. Picturing the other variables in that dimension reveals the intuition that rotating one variable performs a constrained rotation on the others. This is illustrated with the \\code{view\\_manip\\_space()} function."}
view_manip_space2(basis = f_bas, manip_var = f_mvar, lab = colnames(f_dat))
```

@@

```{r step3, echo=F, warning=F, out.width='100%',  fig.cap = "Snapshots of a radial manual tour manipulating aede2: (1) original projection, (2) full contribution, (3) zero contribution, (4) back to original. "}
f_angle <- .29
f_mtour <- manual_tour(f_bas, manip_var = f_mvar, angle = f_angle)
if(dim(f_mtour)[3] != 15) message(
  paste0("!!! n_frames != 15 !!! Check angle step size. n_frames = ", dim(f_mtour)[3], ". angle = ", f_angle))
